---
title: "Forcasting Meal Sales"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Install packages
```{r}
if (!require(dplyr)) install.packages("dplyr")
library(dplyr)
if (!require(forecast)) install.packages("forecast")
library(forecast)
if (!require(TTR)) install.packages("TTR")
library(TTR)
```
Read in data
```{r}
df <- read.csv("meal_orders.csv")
head(df)
```
Inspect Data
```{r}
length(unique(df$meal_id))
length(unique(df$center_id))
nrow(df)
```
This data set has 51 different meal options, 77 different distribution centers, and 456,548 total rows.


Split Data into training and test data sets.  We will use the last 15 weeks as our test data.
```{r}
train <- filter(df, df$week <= 130)
test <- filter(df, df$week > 130)
```


Imagine we work in center 13 and we are interested in forecasting the demand for meal 2290.

```{r}
## Filter for only meal id 2290 and center 13
df_2 = filter(train, train$meal_id == 2290 & train$center_id == 13)
nrow(df_2) #130 rows

#Plot sales over time for center 13 and meal 2290
plot(df_2$week, df_2$num_orders, type = 'l')

## Convert to time-series
df_ts = ts(df_2$num_orders)
plot.ts(df_ts)

## Create test filtered data
test_2 = filter(test, test$meal_id == 2290 & test$center_id == 13)
```
Now we will compare different functions for time series forecasting:

SES() - useful for data w/ no trend or seasonality
```{r}
# Create model. h is the number of periods to forecast
meal_ses = ses(df_2$num_orders, h = 15)
meal_ses$model

# Plot model
autoplot(meal_ses)

# Find Accuracy scores for the model
accuracy(meal_ses, test_2$num_orders)
```
Exponential Smoothing w/ Trend
```{r}
# Create model to predict next 15 weeks
meal_holt = holt(df_2$num_orders, h= 15)
meal_holt$model

# Plot model
autoplot(meal_holt)

# Find Accuracy scores for the model
accuracy(meal_holt, test_2$num_orders)
```

ETS - error, trend, smoothing
```{r}
# Create Model
meal_ets = ets(df_2$num_orders)
autoplot(forecast(meal_ets))
summary(meal_ets)

# Forecast out 15 weeks
checkresiduals(meal_ets)
meal_ets_fore = forecast(meal_ets, h=15)
meal_ets_fore

#Plot
autoplot(meal_ets_fore)

# Score
accuracy(meal_ets_fore, test_2$num_orders)
```

SMA() - simple moving average
```{r}
# Create Model. n is the number of periods to average over
meal_sma = SMA(df_2$num_orders, n=10)

# Plot
plot.ts(meal_sma)

# Score
accuracy(meal_sma, test_2$num_orders)
```

ARIMA:

```{r}
# Create model with differences
meal_ar_diff = diff(df_2$num_orders, differences = 1)
# Plot
plot.ts(meal_ar_diff)

# Create model with auto.arima()
meal_auto = auto.arima(df_2$num_orders)
meal_auto

# Create arima model 0,0,2 from auto.arima optimizer
meal_arima = arima(df_2$num_orders, order = c(0,0,2))
meal_arima
# Forecast out 15 weeks
meal_arima_for = forecast(meal_arima, h=15)
# Plot
autoplot(meal_arima_for)

# Score
accuracy(meal_arima_for, test_2$num_orders)
```
We want to use the model that minimizes our test set error scores from accuracy. For instance, the model that minimizes RSME is ses(), simple exponential smoothing.
Next, you can change around the filters for center id and meal id and see which models work best with which patterns of data.


Additional Resources:
Time series R package list: https://cran.r-project.org/web/views/TimeSeries.html
Package in Python with various time-series models: statsmodels.tsa:  https://www.statsmodels.org/v0.10.0/tsa.html

